Single-cell virus sequencing of influenza infections that trigger innate immunity

### 2018-10-05 paper/paper.tex
Made a few more very minor edits

%%% start IFNrare figure %%%
\begin{figure}
\centerline{
{\bf \Large A}\includegraphics[width=0.25\textwidth,valign=t]{figures/IFN_stochastic/RussellVenn/venn_diagram.pdf}
\hspace{0.02\textwidth}
{\bf \Large B}\includegraphics[width=0.32\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
\hspace{0.02\textwidth}
{\bf \Large C} \includegraphics[width=0.2\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
}
\caption{
IFN expression is rare in influenza-infected cells.
{\bf (A)} Number of infected and IFN+ A549 cells in our prior single-cell transcriptomics~\citep{russell2018extreme}.
\FIGSUPP[IFNrare]{mice} shows that IFN+ cells were also rare in the single-cell transcriptomics of influenza-infected mice by \citet{steuerman2018dissection}.
{\bf (B)} An A549 reporter cell line to identify cells expressing IFN.
The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
We created reporters with type I and type III IFN promoters (\FIGDATA[IFNrare]{reporter_sequences}).
The reporters were efficiently activated by an IFN-inducing strain of Sendai virus (\FIGSUPP[IFNrare]{reporter_validation}).
{\bf (C)}
Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the reporter cell line (see \FIGSUPP[IFNrare]{flow} for details).
The plot also shows uninfected cells, and cells infected with saturating amounts of Sendai virus.
The limit of detection of 0.05\% is indicated with a dashed line, and numbers show the median of three measurements.
}
\label{fig:IFNrare}

@@

\figdata{Sequences of the reporters in \FIG{IFNrare}B are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/plasmids}.}
}
\label{figdata:reporter_sequences}

@@

\subsection{Combined transcriptomics and virus-sequencing of single infected cells}
We next developed a method to determine if features of the infecting virions are responsible for the heterogeneous outcome of infection in single cells.
Standard single-cell transcriptomic techniques are inadequate for this purpose, since they just use 3'-end sequencing of poly-adenylated mRNAs to measure the abundance of each transcript~\citep{klein2015droplet, macosko2015highly, zheng2017massively, cao2017comprehensive, gierahn2017seq}, and do not reveal if the virions that infect individual cells have features (e.g., mutations) that contribute to heterogeneity in infection outcome.

@@

We processed the cells on a commercially available platform~\citep[the 10X Chromium;][]{zheng2017massively} that physically isolates cells in droplets and then reverse transcribes polyadenylated mRNAs to append a unique cell barcode to all cDNAs in each droplet, and a unique molecular identifier (UMI) to each cDNA molecule (\FIG{workflow}B).
Because influenza virus mRNAs are polyadenylated~\citep{robertson1981polyadenylation}, this process appends cell barcodes to both cellular and viral mRNAs.
Furthermore, because virtually the entire influenza genome is transcribed, the cell-barcoded cDNA spans almost all 13,581 nucleotides in the segmented viral genome: the only portions not covered are one universally conserved nucleotide upstream of the transcription start site~\citep{koppstein2015sequencing} and 17 to 22 highly conserved nucleotides downstream of the polyadenylation site~\citep{robertson1981polyadenylation} in each of the eight viral gene segments.

@@

We then called the presence or absence of each viral gene in each infected cell, again using a Poisson model parameterized by background fractions estimated from uninfected canine cells.
Although ten transcripts are encoded by the eight viral genes, we call presence / absence of genes rather than transcripts.
The reason is that the genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable in the 3'-end sequencing transcriptomic data.
\FIG{transcriptomics}D (top panel) shows that the majority (162 of 290) of infected cells express all eight genes, although a substantial minority fail to express one or more genes (see \FIGSUPP[transcriptomics]{frac_has_gene} for frequencies at which individual genes are present).
This measured frequency of infected cells expressing all eight genes is higher than in our own prior work~\citep{russell2018extreme} and studies by others~\citep{brooke2013most, heldt2015single, dou2017analysis}, which have estimated that only 13\% to 50\% of infected cells express all genes. 
The difference is sufficiently modest that it could simply be due to methodological differences in the thresholds used to call gene presence.
But it is also important to remember that the cells in our experiment have been enriched for IFN expression by MACS sorting, which could alter the distribution of viral gene expression from a fully random sample of infected cells.

@@

By examining the synonymous viral barcodes near the 3' termini of transcripts, we determined that 38\% of cells were co-infected with wild-type and synonymously barcoded virions (\FIG{transcriptomics}F; cells called as co-infected if a binomial test rejected null hypothesis that $\ge$95\% of viral mRNA is from one viral barcode variant).
From \FIG{transcriptomics}F, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode (such co-infections cannot be identified from the transcriptomic data).
Interestingly, this co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is a Poisson process.
This discrepancy could arise if the MACS for IFN+ cells also enriches for co-infected cells, if infection itself is not truly Poisson, or if co-infection increases the likelihood that we identify a cell as infected given the thresholds in \FIG{transcriptomics}C.
This moderately high rate of co-infection may also partially explain why relatively more cells in our experiment express all eight viral genes compared to some prior studies, as a co-infecting virion can complement a missing viral gene~\citep{russell2018extreme}.

We next examined expression of IFN and ISGs (\FIG{transcriptomics}G and \FIGSUPP[transcriptomics]{ISG}).
Over 20\% of  infected cells were IFN+ given the heuristic thresholds in \FIG{transcriptomics}C, indicating that the MACS successfully enriched IFN+ cells far beyond their initial frequency of $\sim$0.5\% (\FIG{IFNrare}C) among infected cells.
Few ($\sim$1.3\%) uninfected cells were IFN+; the few that were present might be because the MACS enriched for rare cells that spontaneously activated IFN, or because some cells that we classified as uninfected were actually infected but produced too few viral transcripts to be detected above background.
Many more cells expressed ISGs than IFN itself: the ratio of ISG+ to IFN+ cells was 1.8 among infected cells, and 7 among uninfected cells (\FIGSUPP[transcriptomics]{ISG}A).
The IFN+ cells were a subset of the ISG+ cells: IFN+ cells always expressed ISGs, but many ISG+ cells did not express significant IFN (\FIGSUPP[transcriptomics]{ISG}B).
These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect that they have been infected, but that ISGs are also expressed via paracrine signaling in cells that do not themselves detect viral infection~\citep{stetson2006type,honda2006type}.

Finally, we qualitatively examined how expression of viral genes, IFN, and ISGs relate to the overall structure of the high-dimensional transcriptomic data.
\FIGSUPP[transcriptomics]{tSNE} shows unsupervised t-SNE clustering~\citep{maaten2008visualizing} of the cells.
Although the clustering is entirely unsupervised, cells expressing high levels of viral genes, IFN, and ISGs cluster together---and most of the structure in the t-SNE plot that is not associated with expression of these genes involves uninfected and IFN- cells.

@@

\figsupp[Number of cells for which genotype(s) of infecting viruses were completely determined.]
{Cells for which we could determine the full sequences of all genes expressed by the infecting virus(es).
{\bf (A)} We could call the complete genotypes of the infecting virus(es) for the majority of cells infected with just a single viral barcode variant, but only a minority of cells co-infected with both viral barcodes.
{\bf (B)} The cells for which we could call complete viral genotypes tended to have higher expression of viral mRNA than cells for which we could not call complete genotypes.
Both facts makes sense.
Cells with more viral mRNA are more likely to have their viral cDNA captured in the PacBio sequencing, which is only captures a small fraction of the total transcripts identified by the 3'-end sequencing transcriptomic sequencing.
The lower calling rate for dual-barcode co-infections is probably because these co-infections have more viral genes that must be sequenced (potentially a copy of each viral gene from each viral variant), increasing the chances that one of these genes is missed by the PacBio sequencing. 
An important implication of this plot is that the cells for which we call complete viral genotypes are \emph{not} a random subsampling of all infected cells in the experiment, but are rather enriched for cells that have high levels of viral mRNA and do not have dual-barcode viral infections.
Note also that this plot is limited to the cells that were called as infected (\FIG{transcriptomics}C) and could clearly be classified as IFN- or IFN+ (\FIG{transcriptomics}G).
}
{\includegraphics[width=0.8\textwidth]{figures/single_cell_figures/p_cells_complete.pdf}}
\label{figsupp:ncells}

@@

The foregoing transcriptomic analyses identify which viral genes are expressed in each cell, but not the sequences of these genes.
We therefore used PacBio sequencing (\FIG{workflow}D) to determine the full sequences of the viral genes expressed in the majority of infected cells.

By employing several PCR enrichment strategies (see Methods), we obtained over 200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (\FIGSUPP[genotypes]{CCSs}).
Crucially, the synonymous viral barcodes at both termini of each gene enabled us to estimate the rate of PCR strand exchange (\FIGSUPP[genotypes]{StrandExchange}).
Less then 3\% of the cDNAs experienced strand exchange that we were not able to identify and filter (\FIGSUPP[genotypes]{CCSs}), meaning that the vast majority of CCSs correctly link the sequence of the viral transcript to cell barcodes and UMIs that identify the cell and molecule of origin.

We used these CCSs to call the sequences of viral genes in single cells.
We called a mutation if it was found in least two CCSs originating from different mRNAs (unique UMIs) in that cell, and was also present in at least 30\% of all CCSs for that gene in that cell.
For cells co-infected with both viral barcode variants, we called mutations separately for each viral variant.
This strategy reliably identifies mutations in the virions that initiate infection of cells infected with at most one virion of each viral barcode variant ($\sim$75\% of infected cells), as well as high-abundance mutations in cells co-infected with multiple virions of the same viral barcode.
It will not identify mutations that arise within a cell after the first few rounds of viral genome replication, since such mutations will not reach 30\% frequency in that cell.
Therefore, analogous to somatic variant calling in tumor sequencing~\citep{xu2014comparison, cibulskis2013sensitive}, there is a limit to our detection threshold: we cannot identify mutations that occur on just a small fraction of viral transcripts in a cell.

@@

The 150 cells for which we called the full viral genotypes are shown in \FIG{genotypes}.
Simple visual inspection of this figure reveals a wealth of information.
For instance, the cell with the highest viral burden (\textit{cell 1} in \FIG{genotypes}, which has 65\% of its mRNA from virus) was infected by a virion that expressed wild-type copies all eight genes and did not induce detectable IFN.
But 12 of the other 13 cells with at least 50\% of their mRNA from virus were infected by virions that either had a mutation or failed to express at least one gene, and five of these cells expressed IFN.
The two cells that expressed the most IFN (\textit{cell 13} and \textit{cell 123}) both lacked the viral NS gene, and many other IFN+ cells had different defects such as large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).
However, expressing unmutated copies of all eight genes did not guarantee a favorable outcome for the virus: for instance, the apparently flawless virion that infected \textit{cell 139} only managed to express viral mRNA to 6\% of the total transcriptome, and the apparently flawless virion that infected \textit{cell 105} still induced IFN.

@@

To systematically assess viral features associated with infection outcome, we divided the 150 cells in \FIG{genotypes} into those that expressed unmutated copies of all eight genes (disregarding synonymous mutations) and those that did not (due to gene absence, amino-acid substitution, insertion, or deletion).
\FIG{mutations}A shows that the 49 cells infected by full wild-type viruses had a significantly tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\citep{gini1921measurement}.
Therefore, viral defects are a major contributor to the heterogeneity in viral transcriptional burden among cells.

@@

One other interesting trend emerges from the single-cell data.
There is no difference in the amount of viral mRNA between IFN+ and IFN- cells that express NS---but among cells that lack NS, the IFN+ ones have significantly more viral mRNA (\FIG{mutations}D).
The statistical analysis in \FIG{mutations}D is insufficient to determine if the higher viral transcriptional burden in IFN+ NS-deficient infections is a cause or a consequence of IFN induction, but experiments below will shed more light on this question.

@@

\figsupp[Flow cytometry data for \FIG{validation}A.]
{A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were not infected (three replicates) or infected with stocks of the indicated viral mutant in quadruplicate.
After 13 hours, cells were stained for expression of HA protein and analyzed by FACS for HA and the ZsGreen driven by the \textit{IFNL1} reporter.
The contour plots show the density of all cells, and IFN+ cells are also indicated by orange dots.
Cells were classified as HA+ or IFN+ based on gates set to put 0.05\% of the uninfected cells in these populations.
For all influenza-infected cells, the percentage IFN+ was calculated only among the HA+ cells (since these are the ones that are infected).
For the uninfected cells, the percentage IFN+ was calculated among all cells, since uninfected cells do not express HA.
For each viral mutant, two independently generated stocks were each assayed in duplicate (i.e., \#1a and \#1b are replicates of one viral stock, and \#2a and \#2b are replicates of the other viral stock).
The infections with replicate \#1 of the wild-type virus were performed at an MOI of 0.1 as determined by TCID50, and all other viruses were infected at an equivalent particle number as determined by HI assay. 
The HA staining indicates that the number of transcriptionally active viral units per cell was similar across most mutants, and calculating percentage IFN+ only among HA+ cells largely corrects for modest variation in titer across mutants.
}
{\includegraphics[height=0.75\textheight]{figures/Validation_Figure/SNP_flow_plot.pdf}}
\label{figsupp:SNPflow}

@@

The viral defect most strongly associated with IFN induction in single cells was failure to express the NS gene (\FIG{genotypes}, \FIG{mutations}C).
Although it is sometimes possible to use complementing cells to generate influenza viruses lacking a specific gene~\citep{fujii2003selective,marsh2007specific}, we were unable to generate viruses that lacked NS.
The NS gene encodes two proteins (NS1 and NS2), the first of which is influenza virus's primary innate-immune antagonist~\citep{garcia1998influenza, hale2008multifunctional}.
We therefore mimicked the effect of absence of NS by creating a mutant virus (which we will term ``NS1stop'') that had multiple stop codons early in the NS1 coding sequence.

@@

We tested the rate of IFN induction by each viral stock using the reporter cells.
\FIG{validation} shows that five of the eight mutant viral stocks induced IFN more frequently than wild type.
The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three of the point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced IFN significantly more frequently than wild type.
These results show that the viral defects in single IFN+ cells often cause increased IFN production.

It is tempting to compare the rates of IFN induction measured in \FIG{validation} with the enrichment of particular viral defects among the IFN+ cells in the single-cell data.
However, these comparisons are confounded by the fact that the measurements in \FIG{validation} use \emph{bulk} stocks of each viral mutant---and, as shown by the single-cell data in \FIG{genotypes}, many virions in each stock will also have additional mutations.
For instance, our single-cell experiment used a ``wild-type'' viral stock, but most IFN+ cells (31 of 40) still had viral defects (\FIGSUPP[genotypes]{genotypes_by_ifn}). 
So although increased IFN induction by the mutant viral stocks does indicate the mutations are causally associated with IFN, it is impossible to know how many of the IFN+ cells induced by each stock are infected by virions that also have other defects.
The rate of IFN induction by any given viral mutant can therefore only be accurately quantified by single-cell virus-sequencing.

@@

\section{Discussion}
We have determined the full sequences of all viral genes expressed in single influenza-infected cells, and examined how viral mutations and gene-expression defects associate with infection outcome in each cell.
Methodologically, our major advance is to measure the \emph{genotypes} of viruses in addition to the abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell studies~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, zhu2009growth, schulte2014single, akpinar2016high, heldt2015single, brooke2013most}.
Our method builds on the observations by \citet{saikia2018simultaneous} and \citet{zanini2018virus} that fragmentary viral genetic information can be obtained by more standard single-cell transcriptomic techniques.
To make this information complete, we have coupled a single-cell transcriptomic technique with long-read PacBio sequencing of viral genes, a strategy analogous to that used by \cite{gupta2018single} to obtain full-length isoforms of some cellular genes in single cells.

@@

But although viral variation is clearly important, it still only partially explains heterogeneity among influenza-infected cells.
We find substantial breadth in viral transcriptional burden and occasional IFN induction even among cells infected with fully wild-type virions.
Of course, our single-cell virus sequencing only calls mutations present in a cell at relatively high frequency, so there could be low-frequency mutations that affect cells expressing predominantly wild-type viral genes.
However, we suspect that stochasticity or pre-existing cellular states also affect the outcome of infection, since heterogeneity in IFN induction is present even among cells treated with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell}.

Perhaps the most intriguing question is how the heterogeneity that we have described ultimately affects the macroscopic outcome of infection.
Real human influenza infections are established by just a handful of virions~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza} that then undergo exponential growth, and early IFN responses are amplified by paracrine signaling~\citep{stetson2006type,honda2006type}.
It is therefore plausible that genetic variation in the founding viral population could have a large effect on downstream events.
Extending our approaches to more complex systems could shed further light on how this viral genetic variation interacts with cell-to-cell heterogeneity to shape the race between virus and immune system.

>>>>

%%% start IFNrare figure %%%
\begin{figure}
\centerline{
{\bf \Large A}\includegraphics[width=0.28\textwidth,valign=t]{figures/IFN_stochastic/RussellVenn/venn_diagram.pdf}
\hspace{0.02\textwidth}
{\bf \Large B}\includegraphics[width=0.35\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
\hspace{0.02\textwidth}
{\bf \Large C} \includegraphics[width=0.23\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
}
\caption{
IFN expression is rare in influenza-infected cells.
{\bf (A)} Number of infected and IFN+ A549 cells in our prior single-cell transcriptomics~\citep{russell2018extreme}.
\FIGSUPP[IFNrare]{mice} shows that IFN+ cells were also rare in the single-cell transcriptomics of influenza-infected mice by \citet{steuerman2018dissection}.
{\bf (B)} An A549 reporter cell line to identify cells expressing IFN.
The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
We created reporters with type I and type III IFN promoters (\FIGDATA[IFNrare]{reporter_sequences}).
The reporters were efficiently activated by an IFN-inducing strain of Sendai virus (\FIGSUPP[IFNrare]{reporter_validation}).
{\bf (C)}
Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the reporter cell line (see \FIGSUPP[IFNrare]{flow} for details).
The plot also shows uninfected cells, and cells infected with saturating amounts of Sendai virus.
The limit of detection of 0.05\% is indicated with a dashed line, and numbers show the median of three measurements.
}
\label{fig:IFNrare}

@@

\figdata{Sequences of the reporters in \FIG{IFNrare}B are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/plasmids}.
}
\label{figdata:reporter_sequences}

@@

\subsection{Combined transcriptomics and virus-sequencing of single infected cells}
We next developed a method to determine if features of the infecting virions are responsible for the heterogeneous outcome of infection in single cells.
Standard single-cell transcriptomic techniques are inadequate for this purpose, since they just use 3'-end sequencing of poly-adenylated mRNAs to measure the abundance of each transcript~\citep{klein2015droplet, macosko2015highly, zheng2017massively, cao2017comprehensive, gierahn2017seq}, and do not reveal if the virions that infect individual cells have mutations.

@@

We processed the cells on a commercially available platform~\citep[the 10X Chromium;][]{zheng2017massively} that physically isolates cells in droplets and reverse transcribes polyadenylated mRNAs to append a unique cell barcode to all cDNAs in each droplet, and a unique molecular identifier (UMI) to each cDNA molecule (\FIG{workflow}B).
Because influenza virus mRNAs are polyadenylated~\citep{robertson1981polyadenylation}, this process appends cell barcodes to both cellular and viral mRNAs.
Furthermore, because virtually the entire influenza genome is transcribed, the cell-barcoded cDNA spans almost all 13,581 nucleotides in the segmented viral genome: the only portions not covered are one universally conserved nucleotide upstream of the transcription start site~\citep{koppstein2015sequencing} and 17 to 22 highly conserved nucleotides downstream of the polyadenylation site~\citep{robertson1981polyadenylation} in each of the eight viral gene segments.

@@

We then called the presence or absence of each viral gene in each infected cell, again using a Poisson model parameterized by background fractions estimated from uninfected canine cells.
Although ten transcripts are encoded by the eight viral genes, we call presence / absence of genes rather than transcripts.
The reason is that the genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable by 3'-end sequencing.
\FIG{transcriptomics}D (top panel) shows that most (162 of 290) infected cells express all eight genes, although a substantial minority fail to express one or more genes (see \FIGSUPP[transcriptomics]{frac_has_gene} for frequencies at which individual genes are present).
This measured frequency of infected cells expressing all eight genes is higher than in our own prior work~\citep{russell2018extreme} and studies by others~\citep{brooke2013most, heldt2015single, dou2017analysis}, which have estimated that only 13\% to 50\% of infected cells express all genes. 
However, it is important to remember that the cells in our experiment have been enriched for IFN expression by MACS sorting, which could alter the distribution of viral gene expression from a fully random sample of infected cells.

@@

By examining the synonymous viral barcodes near the 3' termini of transcripts, we determined that 38\% of cells were co-infected with wild-type and synonymously barcoded virions (\FIG{transcriptomics}F; cells called as co-infected if a binomial test rejected null hypothesis that $\ge$95\% of viral mRNA is from one viral barcode variant).
From \FIG{transcriptomics}F, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode (such co-infections cannot be identified from the transcriptomic data).
Interestingly, this co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is Poisson.
This discrepancy could arise if the MACS for IFN+ cells also enriches co-infected cells, if infection itself is not truly Poisson, or if co-infection increases the likelihood that we identify a cell as infected given the thresholds in \FIG{transcriptomics}C.
This moderately high rate of co-infection may also partially explain why more cells in our experiment express all eight viral genes compared to some prior studies, as a co-infecting virion can complement a missing viral gene~\citep{russell2018extreme}.

We next examined expression of IFN and ISGs (\FIG{transcriptomics}G and \FIGSUPP[transcriptomics]{ISG}).
Over 20\% of  infected cells were IFN+ given the heuristic thresholds in \FIG{transcriptomics}C, indicating that the MACS successfully enriched IFN+ cells far beyond their initial frequency of $\sim$0.5\% (\FIG{IFNrare}C) among infected cells.
Few ($\sim$1.3\%) uninfected cells were IFN+; the few that were present might be because the MACS enriched for rare cells that spontaneously activated IFN, or because some cells that we classified as uninfected were actually infected at low levels.
Many more cells expressed ISGs than IFN itself: the ratio of ISG+ to IFN+ cells was 1.8 among infected cells, and 7 among uninfected cells (\FIGSUPP[transcriptomics]{ISG}A).
The IFN+ cells were a subset of the ISG+ cells: IFN+ cells always expressed ISGs, but many ISG+ cells did not express IFN (\FIGSUPP[transcriptomics]{ISG}B).
These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in cells that do not themselves detect infection~\citep{stetson2006type,honda2006type}.

Finally, we qualitatively examined how expression of viral genes, IFN, and ISGs relate to the overall structure of the high-dimensional transcriptomic data.
\FIGSUPP[transcriptomics]{tSNE} shows unsupervised t-SNE clustering~\citep{maaten2008visualizing} of the cells.
Cells expressing high levels of viral genes, IFN, and ISGs cluster together---and most of the structure in the t-SNE plot that is not associated with these genes involves uninfected and IFN- cells.

@@

\figsupp[Number of cells for which genotype(s) of infecting viruses were completely determined.]
{Cells for which we could determine the full sequences of all genes expressed by the infecting virion(s).
{\bf (A)} We could call the complete genotypes of the infecting virion(s) for the majority of cells infected with just a single viral barcode variant, but only a minority of cells co-infected with both viral barcodes.
{\bf (B)} The cells for which we could call complete viral genotypes tended to have higher expression of viral mRNA than cells for which we could not call complete genotypes.
Both facts makes sense.
Cells with more viral mRNA are more likely to have their viral cDNA captured in the PacBio sequencing, which is only captures a small fraction of the total transcripts identified by the 3'-end sequencing transcriptomic sequencing.
The lower calling rate for dual-barcode co-infections is probably because these co-infections have more viral genes that must be sequenced (potentially a copy of each viral gene from each viral variant), increasing the chances that one of these genes is missed by the PacBio sequencing. 
An important implication of this plot is that the cells for which we call complete viral genotypes are \emph{not} a random subsampling of all infected cells in the experiment, but are rather enriched for cells that have high levels of viral mRNA and do not have dual-barcode viral infections.
Note also that this plot is limited to the cells that were called as infected (\FIG{transcriptomics}C) and could clearly be classified as IFN- or IFN+ (\FIG{transcriptomics}G).
}
{\includegraphics[width=0.8\textwidth]{figures/single_cell_figures/p_cells_complete.pdf}}
\label{figsupp:ncells}

@@

We next used PacBio sequencing (\FIG{workflow}D) to determine the full sequences of the viral genes expressed in single infected cells.
Using PCR enrichment (see Methods), we obtained over 200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (\FIGSUPP[genotypes]{CCSs}).
Crucially, the synonymous viral barcodes at both termini of each gene enabled us to estimate the rate of PCR strand exchange (\FIGSUPP[genotypes]{StrandExchange}).
Less then 3\% of the cDNAs experienced strand exchange that we were not able to identify and filter (\FIGSUPP[genotypes]{CCSs}), meaning that the vast majority of CCSs correctly link the sequence of the viral transcript to cell barcodes and UMIs that identify the cell and molecule of origin.

We called a mutation in a viral gene if it was found in least two CCSs originating from different mRNAs (unique UMIs) in that cell, and was also present in at least 30\% of all CCSs for that gene in that cell.
For cells co-infected with both viral barcode variants, we called mutations separately for each viral variant.
This strategy reliably identifies mutations in virions that initiate infection of cells infected with at most one virion of each viral barcode variant ($\sim$75\% of infected cells), as well as high-abundance mutations in cells co-infected with multiple virions of the same viral barcode.
It will not identify mutations that arise within a cell after the first few rounds of viral genome replication, since such mutations will not reach 30\% frequency in that cell.
Therefore, analogous to somatic variant calling in tumor sequencing~\citep{xu2014comparison, cibulskis2013sensitive}, there is a limit to our detection threshold: we cannot identify mutations that occur on just a small fraction of transcripts in a cell.

@@

The 150 cells for which we called the full viral genotypes are shown in \FIG{genotypes}.
Simple visual inspection of this figure reveals a wealth of information.
For instance, the cell with the highest viral burden (\textit{cell 1} in \FIG{genotypes}, which has 65\% of its mRNA from virus) was infected by a virion that expressed wild-type copies all eight genes and did not induce detectable IFN.
But 12 of the other 13 cells with at least 50\% of their mRNA from virus were infected by virions that had a mutation or failed to express at least one gene, and five of these cells expressed IFN.
The two cells that expressed the most IFN (\textit{cell 13} and \textit{cell 123}) both lacked the viral NS gene, and many other IFN+ cells had different defects such as large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).
However, expressing unmutated copies of all eight genes did not guarantee a favorable outcome for the virus: for instance, the apparently flawless virion that infected \textit{cell 139} only managed to express viral mRNA to 6\% of the total transcriptome, and the apparently flawless virion that infected \textit{cell 105} still induced IFN.

@@

To systematically assess viral features associated with infection outcome, we divided the 150 cells in \FIG{genotypes} into those that expressed unmutated copies of all eight genes (disregarding synonymous mutations) and those that did not.
\FIG{mutations}A shows that the 49 cells infected by full wild-type viruses had a significantly tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\citep{gini1921measurement}.
Therefore, viral defects are a major contributor to the heterogeneity in viral transcriptional burden among cells.

@@

One other interesting trend emerges from the single-cell data.
There is no difference in the amount of viral mRNA between IFN+ and IFN- cells that express NS---but among cells that lack NS, the IFN+ ones have significantly more viral mRNA (\FIG{mutations}D).
The statistical analysis in \FIG{mutations}D is insufficient to determine if the higher viral transcriptional burden in IFN+ NS-deficient infections is a cause or a consequence of IFN induction, but experiments below shed more light on this question.

@@

\figsupp[Flow cytometry data for \FIG{validation}A.]
{A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were not infected (three replicates) or infected with stocks of the indicated viral mutant in quadruplicate.
After 13 hours, cells were stained for expression of HA protein and analyzed by FACS for HA and the ZsGreen driven by the \textit{IFNL1} reporter.
The contour plots show the density of all cells, and IFN+ cells are also indicated by orange dots.
Cells were classified as HA+ or IFN+ based on gates set to put 0.05\% of the uninfected cells in these populations.
For all influenza-infected cells, the percentage IFN+ was calculated only among the HA+ cells (since these are the ones that are infected).
For the uninfected cells, the percentage IFN+ was calculated among all cells, since uninfected cells do not express HA.
For each viral mutant, two independently generated stocks were each assayed in duplicate (i.e., \#1a and \#1b are replicates of one viral stock, and \#2a and \#2b are replicates of the other viral stock).
The infections with replicate \#1 of the wild-type virus were performed at an MOI of 0.1 as determined by TCID50, and all other viruses were infected at an equivalent particle number as determined by HI assay. 
The HA staining indicates that the number of transcriptionally active viral units per cell was similar across most mutants, and calculating percentage IFN+ only among HA+ cells largely corrects for modest variation in titer across mutants.
}
{\includegraphics[height=0.71\textheight]{figures/Validation_Figure/SNP_flow_plot.pdf}}
\label{figsupp:SNPflow}

@@

The viral defect most strongly associated with IFN induction was failure to express the NS gene (\FIG{genotypes}, \FIG{mutations}C).
Although it is sometimes possible to use complementing cells to generate influenza viruses lacking a specific gene~\citep{fujii2003selective,marsh2007specific}, we were unable to generate viruses that lacked NS.
The NS gene encodes two proteins (NS1 and NS2), the first of which is influenza's primary innate-immune antagonist~\citep{garcia1998influenza, hale2008multifunctional}.
We therefore mimicked the effect of absence of NS by creating a mutant virus (which we will term ``NS1stop'') that had multiple stop codons early in the NS1 coding sequence.

@@

We tested the rate of IFN induction by each viral stock using the reporter cells.
\FIG{validation} shows that five of the eight mutant viral stocks induced IFN more frequently than wild type.
The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three of the point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced IFN significantly more frequently than wild type.
Therefore, the viral defects in single IFN+ cells often cause increased IFN production.

It is tempting to compare the rates of IFN induction measured in \FIG{validation} with the enrichment of viral defects among the IFN+ cells in the single-cell data.
However, these comparisons are confounded by the fact that the measurements in \FIG{validation} use \emph{bulk} stocks of each viral mutant---and, as shown by the single-cell data in \FIG{genotypes}, many virions in each stock will also have additional mutations.
So although increased IFN induction by the mutant viral stocks does indicate the mutations are causally associated with IFN, it is impossible to know how many of the IFN+ cells induced by each stock are infected by virions that also have other defects.
The rate of IFN induction by any given viral mutant can only be accurately quantified by single-cell virus-sequencing.

@@

\section{Discussion}
We have determined the full sequences of all viral genes expressed in single influenza-infected cells, and examined how viral mutations and gene-expression defects associate with infection outcome in each cell.
Methodologically, our major advance is to measure the \emph{genotypes} of viruses in addition to the abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell studies~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, zhu2009growth, schulte2014single, akpinar2016high, heldt2015single, brooke2013most}.
Our method builds on the observation by \citet{saikia2018simultaneous} and \citet{zanini2018virus} that fragmentary viral genetic information can be obtained by more standard single-cell transcriptomic techniques.
To make this information complete, we have coupled single-cell transcriptomics with long-read PacBio sequencing of viral genes, a strategy analogous to that used by \cite{gupta2018single} to obtain full-length isoforms of some cellular genes in single cells.

@@

But although viral variation is clearly important, it still only partially explains heterogeneity among influenza-infected cells.
We find substantial breadth in viral transcriptional burden and occasional IFN induction even among cells infected with fully wild-type virions.
Of course, our single-cell virus sequencing only calls mutations present in a cell at relatively high frequency, so there could be low-frequency mutations in cells expressing predominantly wild-type viral genes.
However, we suspect that stochasticity or pre-existing cellular states also affect the outcome of infection, since heterogeneity in IFN induction is present even among cells treated with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell}.

Perhaps the most intriguing question is how the heterogeneity that we have described ultimately affects the macroscopic outcome of infection.
Real human influenza infections are established by just a handful of virions~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza} that then undergo exponential growth, and early IFN responses are amplified by paracrine signaling~\citep{stetson2006type,honda2006type}.
It is therefore plausible that genetic variation in the founding viral population could have a large effect on downstream events.
Extending our approaches to more complex systems could shed further light on how viral genetic variation interacts with cell-to-cell heterogeneity to shape the race between virus and immune system.
